Stability and Change in Personal Culture

### 2020-02-10 manuscript9.Rmd
conditional accept revisions

---
title: "Stability and Change in Personal Culture"
author:
- Kevin Kiley
- Stephen Vaisey
date: "Duke University"
output:
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  pdf_document:
    number_sections: yes
header-includes: \usepackage{longtable}
nocite: |
  @swidler01
bibliography: panel_change_bib.bib
abstract: Models of population-wide cultural change tend to invoke one of two broad models of individual change. One approach theorizes that people actively update their beliefs and behaviors in the face of new information. The other argues that, following early socialization experiences, dispositions are stable. We formalize these two models, elaborate empirical implications of each, and derive a simple combined model for comparing them using panel data. We test this model on more than 184 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey. Though the pattern of results is complex, it is somewhat more consistent with the settled dispositions model than the active updating model. Most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes. When persistent change occurs, it is somewhat more likely to occur in younger people than older people and more common for public behaviors and beliefs about high-profile issues than private attitudes. We argue that we need both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
---

@@

Attempts to account for opinion changes in society have produced conflicting theories about the nature of opinion formation and the ability of individuals to maintain consistent attitudes. Some models suggest that people lack the cognitive tools to maintain consistent beliefs on social and political issues. As a result, they construct responses on the fly in the interview setting, drawing on ideas from opinion leaders and changing their attitudes as elite discourse changes [@zaller; @converse; @perrinmcfadden]. By contrast, cohort replacement theories posit that people do hold opinions, are unwilling to alter them in the face of societal change, and thus public opinion changes only with generational turnover [@mannheim; @ryder]. Another set of models claim that people hold "a number of real, stable, and sensible opinions about public policy" and change their opinions in response to new information [@pageshapiro: p. xi; @achen1992]. More recently, sociologists have suggested that people attempt to align malleable peripheral beliefs with relatively fixed "core" beliefs using social cues [@boutylinevaisey; @lakoff; @goldbergstein].

Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to theories in cultural sociology. The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation. This model is rooted in a broadly pragmatist approach to action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001). The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016). These two models represent different approaches to understanding how people come to hold different forms of _personal culture_, or the declarative and non-delcarative attitudes, worldviews, values, dispositions, and associations that manifest at the individual level [@lizardo2016].

In this paper, we make these two models of personal culture change explicit, deduce some of their empirical implications, and derive a statistical model for estimating the prevalence of active updating using panel data. In contrast to previous models that tend to assume one of these data-generating processes to measure stability and reliability of estimates over time, our approach separates persisting change from non-persisting change to estimate if there is evidence of individuals in the population making persistent changes in attitudes and behaviors. We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS). By classifying the pattern of change in personal culture, we can clarify when different accounts of aggregate change are more likely to apply. Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names). We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at bars or attending church, that are important in contemporary U.S. society.

Our analysis yields several results. First, we find that the majority of what appears to be individual-level change in attitude or practice probably reflects short-term (i.e., non-persisting) change or measurement error rather than actual persistent change. Simply put, there is little evidence that large numbers of U.S. adults are changing their beliefs or practices in lasting ways over this period of study. Second, settled opinions vary in how consistently individuals report the same answer. Consistent with theories that argue that people lack clear opinions, some survey items appear to elicit inconsistent or random responses. At the same time, people are consistent on both high- and low-profile public policy items, suggesting a greater degree of "real" attitudes than these theories suggest. Third, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents. On several items it appears that younger adults are still in the process of acquiring dispositions and habits they will take into later life. Fourth, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to be persistent than changes in private attitudes (e.g., political ideology), and people are more likely to report these attitudes consistently. This suggests that interactional and institutional mechanisms may provide stronger support for lasting change than pressures for intrapsychic consistency.

Our findings offer broad support to theories claiming that cultural change comes through generational turnover rather than persuasion and social influence. However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change and that challenge the idea that change in all attitudes follows a similar trajectory over the life course. The pattern of results also supports models of attitude change that put ideological identification at the center of a network of political beliefs and suggest that individuals are more likely to make lasting changes in their partisan identification than to specific policy positions. We argue that there is a place for both the active updating and settled dispositions models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.

@@

How does the respondent formulate a response to that question each time, year after year? To make things as explicit as possible, we can write the data-generating process formally. Although this presentation may make it seem like we are assuming rationality or conscious deliberation, this way of writing the models makes no particular cognitive assumptions at all. We will explain this in greater detail below. For now, consider the following two simple models:

@@

```{r dag, message=FALSE, fig.width=6.8, fig.cap="Different Models of Response Over Time"}
library(ggdag)
library(dagitty)
aum <- 
  dagitty("dag {
        y1 -> y2 -> y3 ;
        v1 -> y1 ;
        v2 -> y2 ;
        v3 -> y3 
        }")
coordinates( aum ) <-
  list( x=c(y1 = 1, y2 = 2, y3 = 3, v1 = 1, v2 = 2, v3 = 3),
        y=c(y1 = 2, y2 = 2, y3 = 2, v1 = 1, v2 = 1, v3 = 1) )

@@

More informally, this AUM posits an agent that is updating her views in the face of social experience. There are formal Bayesian ways of modeling updating, and this model underlies numerous theories of rational updating in the face of new information [@achen1992; @bartelsjackman; @pageshapiro], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply. Following Gross's [-@gross09, pp. 367] pragmatist account, we could instead regard this model as consistent with an "active and creative relation to the world" that "lead[s] actors to see themselves in new ways, to value different kinds of goods, and to become attached to problem solutions that they could not have imagined previously."

The concept of active updating lies at the heart of most theories that suggest social environments influence individual attitudes and behaviors. When sociologists posit that adults adopt the beliefs and behaviors of their friends [@christakisfowler; @centola; @dellapostashimacy], or that individuals adopt behaviors that they view as consistent with their cultural dispositions [@goldbergstein], they invoke this model. When social scientists argue that people adapt their partisan affiliations to conform to their social groupings, change their ideological commitments to conform to their partisan identifications [@greenetal], change their partisan affiliations to match their ideological commitments [@baldassarigelman; @abramowitzsaunders06], or respond to exposure to diverse opinions by becoming more extreme in their viewpoints [@bailetal], the claim is rarely that these changes are temporary shocks that will revert to a stable baseline over time. Instead, the claim is that the changing social environment changes dispositions, which will then remain stable until the environment changes again.

@@

A number of theories of cultural change at the societal level require evidence of active updating at the individual level, such as arguments that the population is becoming more polarized on political issues [@dimaggio_polarization; @iyengarwestwood], arguments that individuals adjust their political preferences based on the performance of the governing party [@pageshapiro; @bartelsjackman], arguments that new ideas diffuse across the population through person-to-person contact [@centola], and arguments that changes in elite discourse drive change in mass opinion [@zaller]. In the absence of evidence that people change their minds in persisting ways, these theories cannot be valid at the aggregate level, and we must seek other explanations for aggregate changes.

@@

The SDM underlies theories that suggest that individuals are stable in their dispositions throughout life or, if they do change, tend to revert back to a relatively stable baseline in short time, including “control” theories such as Identity Control Theory and Affect Control Theory [@robinson2007; @smithlovinheise]. The notion that individuals are stable in their dispositions is commonly associated with cohort replacement theories of attitude change at the population level. These models posit that people hold relatively stable opinions, that few people change their attitudes as they age, and that most change in public opinion must come from older people dying and being replaced in the population with young people who hold different opinions (Mannheim 1952; Ryder 1965). Vaisey and Lizardo (2016), looking at a range of opinion items in the GSS, suggest that change for most likely comes in the form of cohort replacement.

On its face, the settled dispositions model might seem to imply that individuals hold strong, consistent opinions, but this does not need to be the case. The settled dispositions model is also consistent with theories that suggest that people lack stable opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion]. These theories tend to focus on questions of public policy, since the act of opinion creation requires much more information than individuals typically have (Zaller 1992; Converse 1964), and suggest that peoples’ responses at any particular time are a random deviation from a baseline, with this deviation often assumed to be random and commonly referred to as a form of “measurement error” [@pageshapiro; @converse]. As long as these deviations are random, there is no reason to perceive these changes as a form of updating, even if these wave-to-wave differences are quite large.

Like the AUM, the SDM does not require $\nu_{it}$ to be 0 in expectation for the population at time $t$, but it does assume that $\nu_{it}$ has an expected value of 0 within an individual over time. That is, current considerations can move a person from their baseline temporarily, but there is a tendency to "bounce back" to that baseline over time. The SDM thus allows for population-wide shifts in beliefs, practices, or identities at a particular time (temporary period effects), but it assumes that these shifts will be erased over time as people return to their baselines.

@@

The models thus differ fundamentally on their emphasis on the character of personal change. In the AUM, changes tend to persist because they propagate into future responses, shifting the baseline over time. In the SDM, changes tend to revert because each person returns to their existing baseline. We can therefore summarize the models' implications for individual change as follows: the AUM predicts _persisting change_ whereas the SDM predicts _non-persisting change_. We consider the implications of these predictions below.

@@

The clearest theoretical contrast is between models that assume beliefs are settled during adulthood, such as cohort replacement theories, which would produce a pattern of low active updating, and those that posit that individuals update beliefs throughout their lives, which would produce an active updating pattern. This disagreement thus forms the primary structure of our analysis. However, it is possible that some items display active updating while others display settled dispositions. There are at least three ways to understand how both models could be present in the population without contradiction: age-based differences, core-periphery models, and public-private differences.

_Age-based differences:_ The concept of a "cohort effect" is that the environment influences the baseline response tendency when a person is young and then stabilizes for the remainder of one's life [@elder; @bartelsjackman]. To put it differently, cohort formation requires an early period of persistent change followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.

@@

_Core-periphery models:_ A number of theories of attitude development posit that people hold “core” beliefs that they use to adjudicate peripheral attitudes, such as views on specific policy questions [@boutylinevaisey; @goldbergstein]. Much work in political science has sought to understand whether people use ideological identity, policy positions, moral beliefs, or partisanship to adjudicate their position on other issues [@lakoff; @greenetal; @kinderkalmoe]. In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find support for ideological identification as the "core" belief in the political space, consistent with Converse's [-@converse] notion of social constraint. Baldassari and Gelman [-@baldassarigelman] find an increased alignment between partisan identification and policy positions, but little increase in the correlation between specific policy positions over time. They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp. 442]. This suggests that among items tapping issues of politics, most policy issues should show little active updating, but partisan affiliation should demonstrate the highest level.

These core-periphery models suggest that core beliefs should be relatively stable, that changes in core beliefs should necessitate changes in peripheral beliefs, and that people attempt to align their peripheral beliefs with their core beliefs. In our framework, beliefs and behaviors that exhibit stable dispositions are more likely to be “core” beliefs than those that exhibit active updating. Core beliefs should also be more consistently reported, since individuals should know where they stand on these core questions. Determining where one stands on peripheral issues requires greater cognitive effort and therefore are more prone to random perturbations, so should be reported less consistently.

_Public vs. Private Culture:_ A final dimension that likely influences the degree of active updating and consistency of attitude and behavior reports is the publicness of the item, or whether it is in some way externalized in public symbols, discourses, and institutions. The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: pp. 91]. This kind of knowledge is contradictory and flexible, meaning a person can reach and justify a range of different opinions in response to the same prompt, depending on the circumstances. However, when public culture provides a strong "scaffold" through clear signals of how identities and social locations should influence opinions, it becomes easier for people to maintain consistency [@lizardostrand].

Theories that argue that individuals do not hold consistent opinions also posit that issues that receive significant media attention can be reported more consistently than other issues [@zaller; @converse]. This means a small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these. However, we do highlight one issue that saw significant change in elite opinion, which should lead to active updating in the general population. In the time frame we observe (2006-2014), there was a major shift in elite opinion on the issue of gay rights. Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage. For this issue, we expect to find evidence of active updating throughout the population, not just among young people.

Items that tap some public dimension of behavior and attitudes are likely to show higher levels of active updating because social mechanisms, such as joining a group, can help maintain changes in ways that intrapsychic forces cannot. Any item where the burden of maintaining consistency is externalized and not only “in a person’s head” might demonstrate both higher rates of active updating and greater consistency.

@@

First, _to what extent are patterns of cultural change generally better described by an active updating model or a settled dispositions model?_ Previous work on cultural change using cross-sectional data has suggested that cohort effects are generally more important than period effects in explaining broad cultural change [@vaiseylizardo]. This implies that, in the repeated measures data on adults we will use here, we should find that the settled dispositions model performs better on most items because cohort formation should be (mostly) complete.

Second, _is there evidence that younger respondents are doing more active updating than older respondents?_ As we just mentioned, in a sample of adults, the possibility exists that cohort formation may be complete for most beliefs and behaviors before the study period (i.e., before age 18). However, if some cohort formation is still occurring among younger respondents, we should see that evidence consistent with the AUM is disproportionately located among younger respondents.

@@

To investigate these ideas empirically, we examine 184 survey items from the 2006-2014 General Social Survey panels in search of evidence in favor of an active updating model. This period of the GSS contains three different three-wave panels, each of which surveys a sample of adults three times over a four-year period (e.g., 2008-2010-2012). Three waves of data is the minimum amount needed to compare the predictions of the active updating model and the settled dispositions model. We discuss item selection below. While a broader range of years would be preferable, the panel component of the GSS began in 2006 and was discontinued in 2014, so these years represent the full range of what we could analyze using the GSS.

@@

That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response, and that wave 1 carries no additional information in predicting wave 3 once we control for wave 2 (the coefficient on this term should be 0). If change is persisting, then our best guess is that a person's response will be close to what they said last time, and previous responses will provide no additional predictive power.

@@

Since the best estimate of $U_i$ is the mean of the respondent's two previous answers, we can rewrite the SDM prediction like this:

@@

Both of the models above include $y_{i2}$ as a predictor of $y_{i3}$, but only the settled dispositions model includes $y_{i1}$ as a predictor. If the SDM is correct, $y_{i1}$ should be just as predictive as $y_{i2}$ because both are (on average) equally informative about the respondent's stable disposition. Therefore, to test for evidence of active updating we use a model that evaluates whether $y_{i2}$ carries any additional predictive power over $y_{i1}$. If the two previous estimates are equally predictive, then we can be relatively confident that the data we observe came from a settled dispositions model. However, if $y_{i2}$ is a better predictor of $y_{i3}$ than $y_{i1}$ is, we begin to observe evidence that at least some individuals in the population engage in active updating. We use the following non-linear model to estimate the relative influence of $y_{i1}$ and $y_{i2}$:

@@

Rather than estimate separate coefficient estimates for $y_{i2}$ and $y_{i1}$, this model generates two parameter estimates of interest for each item in our analysis: $\beta$, which captures how well any combination of previous waves predicts a person's response at wave 3, and $\phi$, the relative proportion of wave 3 explained by wave 2 compared to wave 1. If the Settled Dispositions Model is the preferred data-generating process for an item, then both $y_{i2}$ and $y_{i1}$ should be equally predictive of $y_{i3}$, and $\phi$ will equal .5, meaning the best estimate of wave 3 is a function of the mean of previous waves, consistent with Equation \@ref(eq:SDM2B). If the active updating model is present in at least some respondents and wave 1 provides no additional predictive power when we control for wave 2, then $\phi$ will increase toward 1 to converge with Equation \@ref(eq:AUM2) in certain circumstances.

>>>>

---
title: "Stability and Change in Personal Culture"
author:
- Kevin Kiley
- Stephen Vaisey
date: "Duke University"
output:
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  pdf_document:
    number_sections: yes
header-includes: \usepackage{longtable}
nocite: |
  @swidler01
bibliography: panel_change_bib.bib
abstract: Models of population-wide cultural change tend to invoke one of two broad models of individual change. One approach theorizes that people actively update their beliefs and behaviors in the face of new information. The other argues that, following early socialization experiences, dispositions are stable. We formalize these two models, elaborate empirical implications of each, and derive a simple combined model for comparing them using panel data. We test this model on 183 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey. Though the pattern of results is complex, it is somewhat more consistent with the settled dispositions model than the active updating model. Most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes. When persistent change occurs, it is somewhat more likely to occur in younger people than older people and more common for public behaviors and beliefs about high-profile issues than private attitudes. We argue that we need both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
---

@@

Attempts to account for opinion changes in society have produced conflicting theories about the process of opinion formation and the ability of individuals to maintain consistent attitudes. Some models suggest that people lack the cognitive tools to maintain consistent beliefs on social and political issues. As a result, they construct responses on the fly in the interview setting, drawing on ideas from opinion leaders and changing their attitudes as elite discourse changes [@zaller; @converse; @perrinmcfadden]. By contrast, cohort replacement theories posit that people do hold opinions and are unwilling to alter them in the face of societal change, and thus public opinion changes only with generational turnover [@mannheim; @ryder]. Another set of models claim that people hold "a number of real, stable, and sensible opinions about public policy" and change their opinions in response to new information [@pageshapiro: p. xi; @achen1992]. More recently, sociologists have suggested that people attempt to align malleable peripheral beliefs with relatively fixed "core" beliefs using social cues [@boutylinevaisey; @lakoff; @goldbergstein].

Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to theories in cultural sociology. The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation. This model is rooted in a broadly pragmatist approach to action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001). The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016). These two models represent different approaches to understanding how people come to hold diverse forms of _personal culture_, or the declarative and non-delcarative attitudes, worldviews, values, dispositions, and associations that manifest at the individual level [@lizardo17].

In this paper, we make these two models of personal culture change explicit, deduce some of their empirical implications, and derive a statistical model for estimating the prevalence of active updating using panel data. In contrast to previous models that tend to assume one of these data-generating processes to measure stability and reliability of estimates over time, our approach separates persisting change from non-persisting change to estimate if there is evidence of individuals in the population making persistent changes in attitudes and behaviors. We then apply this method to 183 items from the 2006-2014 General Social Surveys (GSS). By classifying the pattern of change in personal culture, we can clarify when different accounts of aggregate change are more likely to apply. Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names). We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at bars or attending church, that are important in contemporary U.S. society.

Our analysis yields several results. First, we find that the majority of what appears to be individual-level change in attitude or practice probably reflects short-term (i.e., non-persisting) change or measurement error rather than actual persistent change. Simply put, there is little evidence that large numbers of U.S. adults changed their beliefs or practices in lasting ways over this period of study. Second, settled opinions vary in how consistently individuals report the same answer. Consistent with theories that argue that people lack clear opinions, some survey items appear to elicit inconsistent or random responses. At the same time, people are consistent on both high- and low-profile public policy items, suggesting a greater degree of "real" attitudes than these theories suggest. Third, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents. On several items it appears that younger adults are still in the process of acquiring dispositions and habits they will take into later life. Fourth, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to persist than changes in private attitudes (e.g., political ideology), and people are more likely to report these attitudes and behaviors consistently. This suggests that interactional and institutional mechanisms may provide stronger support for lasting change than pressures for intrapsychic consistency.

Our findings offer broad support to theories claiming that cultural change comes through generational turnover rather than persuasion and social influence. However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change and that challenge the idea that change in all attitudes follows a similar trajectory over the life course. The pattern of results also supports models of attitude change that put ideological identification at the center of a network of political beliefs and suggest that individuals are more likely to make lasting changes in their partisan identification than to their general political beliefs. We argue that there is a place for both the active updating and settled dispositions models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.

@@

How does the respondent formulate a response to that question each time, year after year? To make things as explicit as possible, we can write the data-generating process formally. Although this presentation may make it seem like we are assuming rationality or conscious deliberation, this way of writing the models makes no particular cognitive assumptions. We will explain this in greater detail below. For now, consider the following two simple models:

@@

```{r dag, message=FALSE, fig.width=6.8, fig.cap="Graphical representation of active updating and settled dispositions models over time."}
library(ggdag)
library(dagitty)
aum <- 
  dagitty("dag {
        y1 -> y2 -> y3 ;
        v1 -> y1 ;
        v2 -> y2 ;
        v3 -> y3 
        }")
coordinates( aum ) <-
  list( x=c(y1 = 1, y2 = 2, y3 = 3, v1 = 1, v2 = 2, v3 = 3),
        y=c(y1 = 2, y2 = 2, y3 = 2, v1 = 1, v2 = 1, v3 = 1) )

@@

More informally, this AUM posits a person who is updating her views in the face of social experience. There are formal Bayesian ways of modeling updating, and this model underlies theories of rational updating in the face of new information [@achen1992; @bartelsjackman; @pageshapiro], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply. Following Gross's [-@gross09, pp. 367] pragmatist account, we could instead regard this model as consistent with an "active and creative relation to the world" that "lead[s] actors to see themselves in new ways, to value different kinds of goods, and to become attached to problem solutions that they could not have imagined previously."

Active updating lies at the heart of most theories that suggest social environments influence individual attitudes and behaviors. When sociologists posit that adults adopt the beliefs and behaviors of their friends [@christakisfowler; @centola; @dellapostashimacy], or that individuals adopt behaviors that they view as consistent with their cultural dispositions [@goldbergstein], they invoke this model. When social scientists argue that people adapt their partisan affiliations to conform to their social groupings, change their ideological commitments to conform to their partisan identifications [@greenetal], change their partisan affiliations to match their ideological commitments [@baldassarigelman; @abramowitzsaunders06], or respond to exposure to diverse opinions by becoming more extreme in their viewpoints [@bailetal], the claim is rarely that these changes are temporary shocks that will revert to a stable baseline over time. Instead, the claim is that the changing social environment changes dispositions, which will then remain stable until the environment changes again.

@@

Several theories of cultural change at the societal level require evidence of active updating at the individual level, such as arguments that the individuals in the population are becoming more polarized on political issues [@dimaggio_polarization; @iyengarwestwood], arguments that individuals adjust their political preferences based on the performance of the governing party [@pageshapiro; @bartelsjackman], arguments that new ideas diffuse across the population through person-to-person contact [@centola], and arguments that changes in elite discourse drive change in mass opinion [@zaller]. In the absence of evidence that people change their minds in persisting ways, these theories cannot be valid at the aggregate level, and we must seek other explanations for aggregate changes.

@@

The SDM underlies theories that suggest that individuals are stable in their dispositions throughout life or, if they do change, tend to revert back to a relatively stable baseline in short time, including “control” theories such as Identity Control Theory and Affect Control Theory [@robinson2007; @smithlovinheise]. The notion that individuals are stable in their dispositions is commonly associated with cohort replacement theories of attitude change at the population level. These models posit that people hold relatively stable opinions, that few people change their attitudes as they age, and that most change in public opinion must come from older people dying and being replaced in the population with young people who hold different opinions [@mannheim; @ryder]. Vaisey and Lizardo [-@vaiseylizardo], looking at a range of opinion items in the GSS, suggest that population-wide cultural change most likely comes in the form of cohort replacement.

On its face, the settled dispositions model might seem to imply that individuals hold strong, consistent opinions, but this does not need to be the case. The settled dispositions model is also consistent with theories that suggest that people lack stable opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion]. These theories tend to focus on questions of public policy, since the act of opinion creation requires more information than individuals typically have [@zaller; @converse], and suggest that peoples’ responses at any particular time are a deviation from a baseline, with this deviation often assumed to be random and commonly referred to as a form of “measurement error” [@pageshapiro; @converse]. As long as these deviations are random, there is no reason to perceive these changes as a form of updating, even if wave-to-wave differences are quite large.

Like the AUM, the SDM does not require $\nu_{it}$ to be 0 in expectation for the population at time $t$, but it does assume that $\nu_{it}$ has an expected value of 0 within an individual over time. That is, current considerations can move a person from their baseline temporarily, but there is a tendency to "bounce back" to that baseline over time. The SDM thus allows for population-wide shifts in beliefs, practices, or identities at a particular time (temporary period effects), but it assumes that within people these shifts will be erased over time as people return to their baselines.

@@

The two general models thus differ fundamentally on their emphasis on the character of personal change. In the AUM, changes tend to persist because they propagate into future responses, shifting the baseline over time. In the SDM, changes tend to revert because each person returns to their baseline. We can therefore summarize the models' implications for individual change as follows: the AUM predicts _persisting change_ whereas the SDM predicts _non-persisting change_.

@@

The clearest theoretical contrast is between models that assume beliefs are settled during adulthood, such as cohort replacement theories, which would produce a pattern of low active updating, and those that posit that individuals update beliefs throughout their lives, which would produce an active updating pattern. This disagreement thus forms the primary structure of our analysis. However, it is possible that some items display active updating while others do not. There are at least three ways to understand how both models could be present in the population without contradiction: age-based differences, core-periphery models, and public-private differences.

_Age-based differences:_ The concept of a "cohort effect" is that the environment influences the baseline response tendency when a person is young and then stabilizes for the remainder of one's life [@elder; @bartelsjackman]. To put it differently, cohort formation requires an early period of updating followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.

@@

_Core-periphery models:_ A number of theories of attitude development posit that people hold “core” beliefs, which they use to adjudicate peripheral attitudes [@boutylinevaisey; @goldbergstein]. This work has predominantly focused on political beliefs and adjudicating whether people use ideological identity, policy positions, moral beliefs, or partisanship to form their positions on other issues [@lakoff; @greenetal; @kinderkalmoe]. In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find support for ideological identification -- assessment of oneself as a conservative or a liberal, not as a Republican or Democrat -- as the "core" belief in the political space. Similarly, Baldassari and Gelman [-@baldassarigelman] find increased alignment between partisan identification and policy positions but little increase in the correlation between specific policy positions over time. They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp. 442].

The picture that emerges from these previous works is that ideological identity should be a core, if poorly reported, disposition. Changes in ideological identity should be non-persisting, occurring because people either do not fully understand these terms and report their views with error [@converse] or because people bounce around a stable mean, meaning this item should show low updating and medium to low consistency. On the other hand, we should see evidence that individuals adjust their partisan identification to match their ideological identification as they come to find that one party better represents their core positions [@baldassarigelman]. In other words, liberal Republicans should become liberal Democrats, not conservative Republicans. This means that partisan identification should display high levels of updating. Again, this does not mean that many people change, only that changes tend to persist.

Alongside ideological identificaton, moral issues and general political sentiments might serve as core organizing principles, but what these beliefs are will not be consistent across people [@baldassarigelman]. For example, some people might hold strong, consistent views on abortion while others care about the amount of government intervention in the economy. Because of this heterogeneity, these beliefs should have a moderate level of consistency, since those people who do not value them should report them inconsistently. There should be no evidence of updating in these beliefs, as previous work suggests that people do not bring these attitudes in line with other beliefs, even if they are discordant.

Peripheral beliefs that "grow" out of core beliefs, such as preferences for government spending on various priorities and specific policy positions, should show some evidence of updating, as individuals attempt to align these with their core beliefs. However, because of the cognitive effort required to map peripheral beliefs onto core beliefs, we generally expect the former to be inconsistently reported [@zaller; @converse]. However, there are conditions that might decrease that cognitive effort and therefore increase the amount of updating and consistency with which peripheral issues are reported, to which we now turn.

_Public vs. Private Culture:_ A final dimension that likely influences the degree of active updating and consistency of attitude and behavior reports is the publicness of the item, or whether it is in some way externalized in public symbols, discourses, and institutions. The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: p. 91]. This kind of knowledge is contradictory and flexible, meaning a person can reach and justify a range of different opinions in response to the same prompt, depending on the circumstances. However, when public culture provides a strong "scaffold" through clear signals of how identities and social locations should influence opinions, it becomes easier for people to maintain consistency [@lizardostrand].

Theories that argue that individuals do not hold consistent opinions also posit that issues that receive significant media attention can be reported more consistently than other issues [@zaller; @converse]. This means a small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these, since signals are consistent over time. However, we do highlight one issue that saw significant change in elite opinion, which should have lead to active updating in the general population. In the time frame we observe (2006-2014), there was a major shift in elite opinion on gay rights. Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage. For this issue, we expect to find evidence of active updating throughout the population, not just among young people.

Items that tap some public dimension of behavior and attitudes are also likely to show higher levels of active updating because social mechanisms, such as joining a group, can help maintain changes in ways that intrapsychic forces cannot. Any item where the burden of maintaining consistency is externalized and not only “in a person’s head” might demonstrate both higher rates of active updating and greater consistency.

@@

First, _to what extent are patterns of personal cultural change generally better described by an active updating model or a settled dispositions model?_ Previous work on cultural change using cross-sectional data has suggested that cohort effects are generally more important than period effects in explaining broad cultural change [@vaiseylizardo]. This implies that, in the repeated measures data on adults we will use here, we should find that the settled dispositions model performs better on most items because cohort formation should be (mostly) complete.

Second, _is there evidence that younger respondents are doing more active updating than older respondents?_ In a sample of adults, the possibility exists that cohort formation may be complete for most beliefs and behaviors before the study period (i.e., before age 18). However, if some cohort formation is still occurring among younger respondents, we should see that evidence consistent with the AUM is disproportionately located among younger respondents.

@@

To investigate these ideas empirically, we examine 183 survey items from the 2006-2014 General Social Survey panels in search of evidence in favor of an active updating model. This period of the GSS contains three different three-wave panels, each of which surveys a sample of adults three times over a four-year period (e.g., 2008-2010-2012). Three waves of data is the minimum amount needed to compare the predictions of the active updating model and the settled dispositions model. We discuss item selection below. While a broader range of years would be preferable, the panel component of the GSS began in 2006 and was discontinued in 2014, so these years represent the full range of what we could analyze using the GSS.

@@

That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response, and that wave 1 carries no additional information in predicting wave 3 once we control for wave 2. If change is persisting, then our best guess is that a person's response will be close to what they said last time, and previous responses will provide no additional predictive power.

@@

Since the best estimate of $U_i$ is the mean of the respondent's two previous answers, we can rewrite the SDM prediction as:

@@

Both of the models above include $y_{i2}$ as a predictor of $y_{i3}$, but only the settled dispositions model includes $y_{i1}$ as a predictor. If the SDM is correct, $y_{i1}$ should be just as predictive as $y_{i2}$ because both are (on average) equally informative about the respondent's stable disposition. Therefore, to test for evidence of active updating we use a model that evaluates whether $y_{i2}$ carries any additional predictive power over $y_{i1}$. If the two previous estimates are equally predictive, then we can be relatively confident that the data we observe came from a settled dispositions model. However, if $y_{i2}$ is a better predictor of $y_{i3}$ than $y_{i1}$ is, this provides evidence that some respondents engage in active updating. We use the following non-linear model to estimate the relative influence of $y_{i1}$ and $y_{i2}$:

@@

Rather than generate separate coefficient estimates for $y_{i2}$ and $y_{i1}$, this model generates two parameter estimates of interest: $\beta$, which captures how well any combination of previous waves predicts a person's response at wave 3, and $\phi$, the relative proportion of wave 3 explained by wave 2 compared to wave 1. If the Settled Dispositions Model is the preferred data-generating process for an item, then both $y_{i2}$ and $y_{i1}$ should be equally predictive of $y_{i3}$, and $\phi$ will equal .5, meaning the best estimate of wave 3 is a function of the mean of previous waves, consistent with Equation \@ref(eq:SDM2B). If the active updating model is present in at least some respondents and wave 1 provides no additional predictive power when we control for wave 2, then $\phi$ will increase toward 1 to converge with Equation \@ref(eq:AUM2) in certain circumstances.

Our estimates of $\beta$ provide a measure of the consistency of individuals' responses, contingent on the degree of active updating. We can think of this parameter as analogous to an $R^2$ measure in a traditional linear model, capturing the total "predictiveness" of the model. If individuals pick a random response at each wave, the best predictor for a person at wave 3 will be the sample average, and $\beta = 0$. If there is little random fluctuation between waves, once the amount of active updating is accounted for,  $\beta$ will approach 1.
